---
title: "Quantifying Levallois core shape using 3D geometric morphometrics: a new method applied to Nubian cores"
author:
  - Emily Hallinan:
      correspondence: "yes"
      email: eshallinan@ualg.pt
      orcid: 0000-0002-0698-8238
      institute: ICArEHB
  - João Cascalheira:
      institute: ICArEHB
      orcid: 0000-0003-0321-8892
institute:
  - fosg:
      name: Interdiscipinary Center for Archaeology and Evolution of Human Behavior
      address: Universidade do Algarve, Campus de Gambelas, 8005-139 Faro, Portugal
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
abstract: |
  Text of abstract
keywords: |
  3D geometric morphometrics; Levallois; Lithic technology; Nubian Levallois cores
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
```


# 1. Introduction

Levallois technology is regarded as a hallmark of the Middle Stone Age (MSA) in Africa and the Middle Palaeolithic of Europe and western Asia, associated with at least three hominin species – archaic humans, anatomically modern humans and Neanderthals (Foley and Lahr 1997; Tryon et al. 2005; Hublin 2009). Named after the area in France, Levallois-Perret, where it was first described (Boucher de Perthes 1857; de Mortillet 1883; Commont 1909), Levallois is a type of prepared core technology that involves preparation of the convexities of the core surface to control the shape and size of the final end-product. The degree of planning involved in the pretermination of the Levallois end-product has been viewed as an indicator of cognitive complexity owing to the conceptualisation of the final core and product form, the sequential method required to achieve it, and social learning to transmit the process (Schlanger 1996; Lahr and Foley 2001; Wynn and Coolidge 2004; Lycett et al. 2016). The implications of Levallois for technical and economic behaviour has also seen varied interpretations, related to its efficiency (Brantingham and Kuhn 2001; Sandgathe 2004; Lycett and Eren 2013) and the role of its end-products (Sisk and Shea 2009; Eren and Lycett 2016; Shimelmitz and Kuhn 2018). Despite the central place that Levallois occupies in human evolution, its definition (Copeland 1983; Van Peer 1992; Dibble and Bar-Yosef 1995; Chazan 1997) and identification of both the technology (Hu et al. 2019a, 2019b; Li et al. 2019; Pallo 2022) and specific Levallois methods in assemblages (Rose et al. 2011; Goder-Goldberger et al. 2016; Blinkhorn et al. 2021a, 2021b, 2022; Hallinan et al. 2022a) remain highly debated topics.

While definitions of Levallois have shifted over the past decades, shape remains a key concept in all of them. Bordes (1950, 1961, 1980) emphasised the predetermined shape of the Levallois flake through special preparation of the core prior to its removal. Different configurations of preparatory removals from the core surface would produce end-products with different morphologies – flakes, blades or points. The classic Levallois ‘tortoise’ flake core was ovate with a flat upper surface that was shaped by a series of centripetal removals from around the perimeter, creating a form that resembled a tortoise carapace. The central preferential Levallois flake was then struck from a prepared striking platform on the perimeter, producing a large, flat oval blank. For Bordes, the morphology of the Levallois core was strongly tied to the end-product shape: an oval core produced an oval flake, a triangular core produced a triangular flake.

Subsequent technological approaches have shifted attention away from the final flake form towards understanding the Levallois reduction process (Bar-Yosef and Dibble 1995). The most widely applied definition is that of Boëda (1988, 1994, 1995), focusing on the geometric structure of the Levallois core that is conceptualised as a volume and worked through a series of steps. Under Boëda’s Levallois concept, a set of criteria must be fulfilled. The volume of a core consists of two asymmetric convex hemispheres that possess a plane of intersection at the core’s margin. These hemispheres or surfaces are hierarchically related, each serving a specific, fixed role in reduction: a preparation surface for the production of striking platforms, and a flaking surface for the removal of the Levallois product. The flaking surface possesses both lateral and distal convexities that must be maintained to control the direction of force for the end-product. The Levallois product is removed parallel to the plane of intersection, using direct percussion with a hard hammer. Different Levallois strategies are distinguished whereby predetermined end-products can either be preferential (producing a single blank) or recurrent (producing multiple successive blanks), prior to the repreparation of the flaking surface. Viewed within a technological framework, it is the pattern and orientation of the preparatory flaking – not the core shape itself – that determines the morphology of the end-product (Boëda 1994, 1995; Van Peer 1992).

# 2. Geometric Morphometrics and Levallois morphology

### 2.1. Principles of shape in GM and Levallois technology

Given that shape is so integral to the Levallois concept, it is perhaps surprising that geometric morphometric techniques have seldom been applied to questions related to its definition. Geometric morphometric (GM) approaches provide a statistical framework for studying shape variation – where shape is the specific geometric configuration of a specimen – independently of size (Slice 2007). Originally developed for applications in biological sciences (Rohlf and Marcus 1993; Adams et al. 2004), GM has gained traction as a powerful tool for quantitative analysis of shape across different aspects of lithic studies, though most commonly applied to bifaces, flakes and points since these can be consistently orientated and aligned according to geometrically correspondent points (e.g. Lycett et al. 2006; Archer and Braun 2010; Iovita 2011; Archer et al. 2016, 2018; Herzlinger et al. 2017; Herzlinger and Grosman 2018; Archer and Presynakova 2019; Okumura and Araujo 2019; Timbrell et al. 2022a). As a consequence, most current GM studies that relate to Levallois technology have focused on debitage, using two-dimensional (Eren and Lycett 2012; Picin et al. 2014; Buchanan et al. 2023) and, occasionally, three-dimensional techniques (Chaćon et al. 2016; Bustos-Perez et al. 2024). The problem of identifying homologous points between specimens for comparison has meant that the application of GM to cores has been limited, since this artefact type displays marked variability in form through continuous, non-uniform reduction trajectories. However, the highly structured geometry of Levallois cores means that common features can be identified between specimens and therefore shape can be compared objectively (Lycett et al. 2010; Lycett and von Cramon-Taubadel 2013).

Elsewhere, 3D methods have been used to test the definition of Levallois cores against that of discoidal cores according to Boëda’s (1993, 1994) criteria, aiming to assess whether the latter can be regarded as a ‘formal’ (as opposed to expedient) technology (Ranhorn et al. 2019). This compared the volumes of the hemispheres and angles of the flaking surfaces – features that are central to distinguishing the two technologies, and that are extremely difficult to quantify using traditional measurement methods. While landmarks were used to delineate different features of the core (i.e., the plane of intersection between the two hemispheres, the outline of the main flake removal), this analysis did not explicitly use GM techniques (Ranhorn et al. 2019).

In the only prior work addressing shape variability in preferential Levallois cores, 3D GM was applied to preferential centripetal Levallois cores to assess relative levels of shape variation across the core surfaces in specimens from Africa and Eurasia (Lycett and von Cramon-Taubadel 2013). This found relatively limited differences between core samples across space and time, with outline (planform) shape showing the greatest variability, whereas the least variation was seen in the topology of the upper flaking surface. This is consistent with the Levallois concept which concerns controlling the convexity of the upper surface. Critically, this also implies that outline shape of Levallois cores is the attribute most likely to reflect regionally or temporally distinct traditions, with Nubian Levallois cited as one possible example of this (Lycett and von Cramon-Taubadel 2013). Nubian Levallois was also central to an earlier study of the factors affecting Levallois end-product shape in Van Peer’s (1992) analysis of Nile Valley Middle Palaeolithic assemblages. Through attribute analysis and refitting, Van Peer demonstrated that, rather than core shape, the organisation of scar ridges on the Levallois surface are key in guiding the preferential removal and thus are what dictate the end-product shape. In Nubian cores, the steep angle of this guiding distal median ridge (DMR) is regarded as a characteristic feature of Nubian core morphology (Usik et al. 2013).

Nubian Levallois describes a specific method for producing pointed end-products, named after the Nubian region of present-day southern Egypt and northern Sudan where it was first defined (Guichard and Guichard 1965; Hallinan and Marks 2023). Recently, the strict definition of Nubian Levallois cores has become important for identifying cores in regions beyond Nubia, with implications for the distribution of Nubian technology in terms of human demographic and cultural behaviour (Usik et al. 2013; Hallinan et al. 2022b). Shape is an important variable in a number of the defining attributes of Nubian technology and its end-products, but currently this is recorded descriptively (e.g. triangular, cordiform, pitched; Usik et al. 2013), or characterised using indices of elongation (length to width ratio), and flattening (width to thickness) (e.g. Blinkhorn et al. 2021a; Samawi and Hallinan 2024). However, this does not adequately capture shape either in two-dimensional outline or three-dimensional volume, and therefore important aspects of Nubian Levallois morphology and variability remain unexplored.

This paper presents a new methodology for using 3D digital techniques to characterise the attributes that define Nubian Levallois cores. First, we outline the steps used to extract the relevant data from 3D models of scanned artefacts, providing R script and data to replicate the process in full. Then, we apply our method to 3D analysis of Middle Palaeolithic assemblages from sites with Nubian Levallois cores from two regions: the Nile Valley in Egypt, and Dhofar in southern Oman. Using the analytical tools of a GM approach, we address morphological variability within and between Nubian core assemblages and test a series of hypotheses arising out of prior work on shape in Levallois technology.

### 2.2. Core shape and size

All prior comparative studies of Nubian Levallois cores have used core size as the primary quantitative basis for analysis (e.g. Hilbert et al. 2016; Groucutt and Rose 2023; Samawi and Hallinan 2024). Size is important in studies of lithic variability and human technological and economic behaviour because stone knapping is a reductive process, and artefacts are observed in their discarded state (Dibble et al. 2017). For certain lithic artefact forms, such as bifaces (i.e. handaxes, bifacially-retouched points; Archer and Braun 2010; Key and Gowlett 2023; Thulman et al. 2023), artefact shape is demonstrated to change as they are progressively reduced in size through maintenance and resharpening. However, a characteristic of Levallois reduction is “auto-correlation” (Boëda 2013: 158), whereby the shape of the core will be maintained throughout the knapping process, regardless of its phase of reduction. Here, we test the expectation that the overall Nubian core shape template will be preserved in spite of size differences due to progressive reduction due to the strong geometric principles structuring Levallois technology.

### 2.3. Core preparation strategy

Van Peer (1992) proposed that the organisation of scar ridges on the Levallois surface are key in guiding the preferential removal and thus are what dictate the end-product shape. Nubian Levallois is a distinctive strategy of core preparation that creates a specific pointed morphology. We test whether variation in the preparation method (‘Type 1’ distal, ‘Type 2’ lateral, and ‘Type 1/2’ distal and lateral) affects core and end-product shape.



### 2.4. End-product shape
Bordes (1950, 1980) emphasised that the morphology of the Levallois core was strongly tied to the end-product shape. We test this by comparing core outline shape with the shape of preferential removal scars. For Nile Valley assemblages, where both cores and end-products are present, we further test whether the final preferential scar shape on discarded cores is representative of the shape of Levallois products in the same assemblage.

### 2.5. Regional traditions
GM analyses by Lycett and von Cramon-Taubadel (2013) indicate that the outline shape of Levallois cores, as opposed to topology of the two surfaces, is most likely to reflect regionally or temporally distinct traditions. We test this by comparing three-dimensional aspects of core shape (planform outline and surface convexities) between Nubian core samples from two regions, separated by 2500 km.

# 3. Materials and methods

### 3.1. Materials
The lithic samples derive from the Nile Valley sites of Nazlet Khater 1 (Middle and Upper levels) and Nazlet Khater 3 (NK), and Dhofar sites of TH.571 and TH.584 (TH). These assemblages contain Nubian Levallois cores whose technological characterisation is not disputed, meeting the generally-accepted criteria for this artefact type (Usik et al. 2013; Hallinan et al. 2022b). Nubian reduction is characterised by three main strategies of preparation to create the central ridge that guides the preferential removal. This can be achieved through distal removals (elsewhere described as ‘Type 1 Nubian’ (Guichard and Guichard 1965)), lateral removals (‘Type 2’), and a combination of both distal and lateral removals (‘Type 1/2’ (Olszewski et al. 2010; Usik et al. 2013)). The analysed samples in this study include only complete cores which preserve a preferential scar outline and a clear DMR (Table 1, Table 2). Levallois end-products from NK were included for comparison with the scar outline. These possessed varied dorsal scar patterns, meaning that they could derive from both Nubian or ‘Classical’ Levallois cores (Van Peer 1992). Products in the analysed sample had a generally convergent distal end shape, although this encompassed products with wide-ranging morphologies (Table 3).

```{r}

assemblages <- read_csv("../data/raw_data/assemblages.csv")

knitr::kable(assemblages, caption = "Table 1: Summary of assemblage samples")

```

```{r}
preparation_strategies <- read_csv("../data/raw_data/preparation_strategies.csv")

knitr::kable(preparation_strategies, caption = "Table 2: Nubian core preparation strategies in NK and TH assemblages")
```

```{r}

```

#### 3.1.1. Nazlet Khater (NK)
The NK sites were identified during surveys on the west bank of the River Nile in 1976 as part of the Belgium Middle Egypt Prehistoric Project, with the lithic material currently curated at Leuven University, Belgium (Vermeersch et al. 1977, 1979). The studied assemblages derive from buried – albeit undated – stratified contexts, at NK1 divided into Middle (NK1-M) and Upper (NK1-U) assemblages. The underlying Lower assemblage was excluded from this analysis owing to the small sample size, and although the Upper layer was initially divided into A and B horizons, these were later deemed to be homogenous by the excavators (Vermeersch et al. 2002a). Both NK1 and NK3 were interpreted as chert quarrying sites owing to the availability of high-quality chert cobbles in nearby Nile and wadi deposits (Vermeersch et al. 2002a, 2002b). The Nubian Levallois reduction method occurs alongside ‘Classical’ centripetally-prepared Levallois flake production. While Nubian cores substantially outnumber ‘Classical’ cores (the latter comprising 25% and 18% of the Levallois cores in the Middle and Upper assemblages respectively), these proportions do not correspond with Levallois end-products attributed to each method based on dorsal scar patterns (‘Classical’: NK1-M 55%; NK1-U 49%) (Vermeersch et al. 2002a, 2002b). This raises two questions; firstly, whether end-products can be reliably distinguished from different Levallois reduction methods, and secondly, whether during a full reduction sequence the two methods could be applied to the same cores. Van Peer (1992) rejects this latter scenario of flexible implementation of different Levallois methods since it is unsupported by the refit sequences. However, in a brief reassessment of 35 ‘non-Nubian’ Levallois cores from NK1-Upper (presumed to be categorised as Classical in previous analyses), numerous cores were observed to possess a remnant distal platform and muted distal median ridge, suggesting there may be some more nuanced aspects to Levallois reduction at the sites. 

The 115 Nubian cores included in the sample show a predominantly distal preparation strategy (75.7%), with a small number of bilaterally prepared cores (6.1%), and in contrast to previous analyses of the assemblages, Nubian cores with both distal and lateral removals were also distinguished (16.5%) (Table 2).


#### 3.1.2 Dhofar (TH)
The TH assemblages were collected in 2023 as part of the ongoing Dhofar Archaeological Project which has identified Nubian Levallois technology at hundreds of localities on the interior Nejd plateau near Thumrait in southern Oman (Rose et al. 2011; Rose et al. 2023). The samples TH.571 and TH.584 derive from surface contexts where artefacts lie on a desert pavement surface, but an OSL age of 106 ka was obtained from a buried core at Aybut Al Auwal (TH.59), a few hundred metres from TH.584 (Rose et al. 2011). Chert is available from primary outcrops close to each site.

There are several key differences between the TH and NK assemblages: first, Nubian Levallois is the only Levallois core reduction strategy present in almost all Dhofar Middle Palaeolithic assemblages, and furthermore there are very few non-Levallois cores. Additionally, Levallois end-products are scarce at most sites which may reflect different patterns of landscape use between the Nile Valley and the Nejd plateau, as well as the types of site studied. The Dhofar Nubian cores also show more varied patterns of preparation than the predominantly distal preparation at NK, using mainly bilateral (41.2%) and a combination of distal and lateral (45.1%) removals (Table 2). It is worth noting that at other Dhofar sites, distally-prepared cores are proportionally more common than in these two particular assemblages (Usik et al. 2013; Rose et al. 2018).

### 3.2. Methods
The specific features that define Nubian cores were captured on 3D models of artefacts using landmark and semilandmark points, complemented by qualitative attribute data (Table 4). Landmark processing, GM and statistical analyses were carried out in R (R Core Team 2024) using functions in the packages Morpho (Schlager 2017) and geomorph (Baken et al. 2021; Adams et al. 2024). Full details are provided in the accompanying R markdown and R scripts (REPOSITORY), and extended results (Supplementary figures and tables).

```{r}

```


#### 3.2.1. 3D data acquisition
Artefacts were scanned by one of us (EH) using an Einscan Pro 2x Plus structured light scanner. Each specimen was captured in multiple orientations to ensure complete coverage and these scans were automatically aligned in the associated Shining 3D software. Mesh models were also generated with this software at medium resolution, exported in PLY format, and simplified using the quadric edge decimation resampling tool in Meshlab (quality threshold 0.3).

#### 3.2.3. Landmarking: cores
Defining landmarks as homologous points that correspond between specimens is complex on archaeological artefacts and therefore these must be chosen carefully and consistently. A new protocol for landmarking cores was adapted from the method used by Archer et al. (2018, 2021) for blanks, using the software Landmark V3 (University of California, Davis) for landmark placement with subsequent processing in R (R Core Team 2024). Technologically meaningful, homologous points were identified based on defining morphological features of Nubian cores: the opposed proximal (preferential) and distal (preparation) platforms. One landmark was placed at the distal extremity and two at the lateral extremities of the proximal platform. Three curves consisting of 20 semilandmarks (SLMs) each were placed around the core edge between these points, bisecting the upper and lower surfaces of the Levallois core (Figure 1a). Seven SLMs were placed along the DMR in order to take digital angle measurements; however, they are excluded from the analysis presented here due to complicated measurement parameters (Valletta et al. 2020; Yezzi-Woodley et al. 2021; Schunk et al. 2023; Muller et al. 2024). The preferential (or final) scar was outlined with 10 platform and 40 edge SLMs (Figure 1b). Two surface patches were placed on the upper and lower core faces of one arbitrarily selected artefact to be used as a template for other specimens (Figure 1c)

These co-ordinate configurations were imported into R with the platform points treated as fixed landmarks and the points around the core and preferential scar outline as sliding SLMs. These were resampled as equally-spaced SLMs and then slid along the mesh model surface (Figure 1d-e). The SLM configuration of the surface-patched specimen was deformed onto the other specimens to generate geometrically correspondent surface co-ordinates across all cores.

#### 3.2.2. Landmarking: end-products (blanks)
The landmarking protocol followed by Archer et al. (2018, 2021) for blanks was replicated on Levallois end-products. Three fixed landmarks were placed on the platform (at the point of percussion and lateral extremities of the platform) to allow uniform orientation of the artefacts. Four curves formed of SLMs between these points (three platform curves of 5, 5 and 10 SLMs, and one of 40 around the flake edge) served to capture the flake outline (Figure 2a). Three surface patches were placed on the platform, dorsal and ventral faces of one arbitrarily selected template artefact (Figure 2b). Following the same steps as for cores outlined above, edge curve SLMs were resampled and slid (Figure 2c), and the surface-patched template was deformed across all of the other specimens. Lastly, the SLMs outlining the ventral face were extracted and resampled for shape comparison with core preferential scar outlines (Figure 2d).


```{r}

```


```{r}

```

#### 3.2.4. GM analyses
The processed landmark configurations were subjected to Procrustes superimposition, a process required for statistical shape analysis that removes differences related to size and orientation by scaling and aligning all specimens (Rohlf and Corti 2000). Generalised Procrustes Analysis (GPA) superimposes landmark configurations using the mean shape as the target for comparison (Adams and Otárola-Castillo 2013). Subsequent statistical analyses were carried out on the co-ordinates of these Procrustes-aligned shapes.

To assess allometry (i.e. whether size and shape vary independently), artefact shapes were regressed against their log-transformed centroid sizes using a generalised linear model (Klingenberg 2016). In GM, centroid size is a measure of overall size that treats size as mathematically independent of shape (Zelditch et al. 2012), therefore does not rely on the distance between any two points (Bookstein 1991). It is calculated as the square root of the sum of squared Euclidean distances of all landmarks from their centroid (the mean of all landmark coordinates; the centre of gravity). The regression slope of the (Type 1) linear model was compared against a null model of isometric change in a Procrustes ANOVA on the sum of squared Procrustes distances (Goodall 1991). In the null model, isometric change (size and shape are independent) has a slope of zero, and allometric change (size and shape are dependent) a non-zero slope. The significance of covariation of shape and size was tested using a non-parametric residual randomised permutation procedure with 1000 permutations, based on Goodall’s F-statistic (Goodall 1991).

Principal Components Analysis (PCA) was performed on the covariance matrix of the aligned artefact shapes, with each component capturing aspects of variation in shape space. Outliers were identified as specimens with Procrustes distances exceeding three standard deviations of the mean shape and were excluded from the PCA (Supplementary Table S1). PC plots of the components accounting for the greatest variability in the dataset allowed comparison of groups in the data (core surfaces, preferential scars and NK products), using attributes such as region and core preparation strategy. The minimum and maximum extremes of shape on each PC axis were visualised in relation to the mean landmark configuration in order to understand the nature of shape variation each component described. One-way ANOVAs using Type III sums of squares were conducted on individual PC scores to test for significant differences between groups. Further multivariate ANOVAs (MANOVA) were used to examine interactions between variables where relevant. To understand the specific differences between pairs of attributes, post-hoc Tukey’s Honest Significant Difference (HSD) tests compared the means of each combination of categories.

To compare covariation between two sets of co-ordinates on the same specimen (core outline shape and preferential scar shape), a two-block partial least squares method (PLS) was used (Rohlf and Corti 2000). This relationship was visualised by plotting two-block PLS scores of core outlines against those of preferential scars on the first dimension (i.e. the maximum covariance) using a generalised linear model.


# 4. Results
### 4.1 Shape and size
In order to examine the relationship between shape and artefact size in the Nubian core assemblages, first standard artefact dimensions for each assemblage are presented, then GM analyses based on centroid size are carried out. Existing metric data were available for the whole NK assemblages (Van Peer 1992; Vermeersch et al. 2002a, 2002b), while new measurements were recorded for the TH assemblages. Comparison of the two regional samples show the mean length of TH cores is nearly twice that of NK cores (TH mean: 111 mm, NK mean: 60 mm) (Table 5, Table 6). Furthermore, even the smallest TH cores are larger than the NK average (TH range: 68.2-174.5 mm). Although metric variability is limited between the three NK assemblages, TH.571 is markedly larger than TH.584, by a factor of 1.3 across all dimensions. Elongation (length/width) and flattening (width/thickness) indices indicate that TH cores are more elongated (1.4) than NK (1.2), while NK1 cores are flatter (2.6) than those at NK3 and the TH sites (2.4). Therefore, the principal differences between regional assemblages are that TH cores are both substantially larger, and are more elongated, than NK cores.



When centroid size is used as an overall indicator of core size, this upholds the same patterns as length, whereby the mean value for TH cores is also double that of NK (TH: 835, NK: 443) Figure 3, Supplementary Table S2). These differences were found to be strongly statistically significant in a Welch’s T-test (t = 19.8, p < 0.001). Similarly, there is limited variation in mean centroid size among the NK assemblages (NK1-M: 455, NK1-U: 436, NK3: 445), while TH.571 (892) is larger than TH.584 (684) by a factor of 1.3, as observed in the dimensional metrics. Using the coefficient of variation (CV, calculated as the ratio of the standard deviation to the mean, expressed as a percentage) to assess size standardisation (Eerkens and Bettinger 2001), all assemblages show very high standardisation; NK3 shows the lowest (21%), and TH.571 the highest (10%) (Supplementary Table S2). Regional mean values indicate higher standardisation among Dhofar cores (16%) than the Nile Valley (17%), which is similar to observations of CV on length for other assemblages from Dhofar (16%) and the Nile Valley (22%) (Samawi and Hallinan 2024). The overall strong agreement with standard metric results demonstrates that centroid size can be a useful substitute in place of taking standard dimensional measurements on specimens that are subject to 3D scanning and GM analysis.



Allometry was assessed for core surfaces (i.e. 3D shape), core outlines (simplified to 2D planform shape), and preferential scar outline, regressing these shapes against the natural log of their centroid sizes using a generalised linear model (Figure 4). The regression for all shapes shows a positive correlation, with distinct regional clusters of NK and TH cores. A Procrustes ANOVA for the whole Nubian core sample (i.e. not separating samples regionally) shows statistically significant weak to moderate positive allometry across all shapes (core surfaces: F = 11.00, p < 0.001; core outlines: F = 16.65, p < 0.001), being strongest for preferential scars (F = 33.26, p < 0.001) (Table 7, Supplementary Table S3). However, given the marked centroid size differences between TH and NK cores, separate Procrustes ANOVAs were carried out to test whether region was the driver behind (albeit weak) positive allometry. The results for core surfaces show much lower F-statistic values, with weaker statistical significance (NK: F = 2.14, p < 0.05; TH: F = 1.60, p < 0.1), indicating that centroid size is a limited explanatory factor for intra-regional shape variation in Nubian cores (NK: 1.9% of variation; TH: 3.2%) (Table 7). Similarly low, though statistically significant, F-values were observed for preferential scars on cores (Table 7), favouring the null hypothesis that Nubian core shape variation is independent of differences in size in each region. Overall, when regional differences are removed, size is not a significant factor in determining Nubian core shape variation. This is consistent with a Levallois strategy that focuses on preparing a core surface to produce an end-product of a specific shape. Furthermore, it suggests that there may be other behavioural factors related to raw material, environment, function or culture that explain inter-regional core size differences (Samawi and Hallinan 2024). 




### 4.2. Shape variability
A PCA was conducted to explore aspects of Nubian core shape variation independent of size. The different principal components (PC) capture both the nature and percentage of total shape variance, showing that for all core surfaces, PC1 explains 19.7% of variance, followed by PC2 (14.9%) and PC3 (11.2%), with 66% of variance accounted for by the first six PCs (Supplementary Table S4). PC1 mainly explains elongation, with high values indicating short, wide cores, while low values correspond to long, narrow cores (Figure 5). High PC1 values also indicate a steep distal ridge angle, whereas this angle is shallow for low values. PC2 is driven by lateral skew, relating to the orientation of the distal tip and lateral twist in the profile of the plane of intersection. PC3 relates to core profile and the distribution of mass between the two flaking surfaces (Figure 6). Cores with high PC3 scores have greater volume concentrated in their upper surface, whereas the lower surface volume is greater for low scores. Additionally, cores with high PC3 values possess a steeper distal ridge, though this may partly be explained by the increased convexity of the upper face. Likewise, low PC3 scores correspond to a shallower distal convexity. PC4 describes the pointedness of the distal end of the core; low scores indicate a triangular shape, while high scores are more ovate with a rounded distal. PCs 5 and 6 account for only 6-7% of the variance each but capture important differences in core thickness and convexity. High PC5 scores reflect thicker cores with pronounced surface convexities and a steep distal ridge, with low scores indicating flatter surfaces and a muted distal ridge angle. PC6 captures core thickness, particularly in the volume of the lower surface. Cores with low PC6 scores are flat with limited lower surface volume, while cores with high scores are thick with a large lower surface volume. 


#### 4.2.1. Regional variability
ANOVAs and appropriate post-hoc tests were performed on the key variables associated with the specimens: region, assemblage, preparation strategy and Nubian core type (Table 8, Supplementary Table S5). High F-statistics indicate more variability between, than within groups. The strongest shape differences are seen inter-regionally between the Nile Valley (NK) and Dhofar (TH), particularly on PC1 (F = 44.97) and PC3 (F = 39.84, p <0.001), with moderate strength on PC2 (F = 8.33) and PC4 (F = 9.29, p <0.005), and weak on PC5 (F = 6.20) and PC6 (F = 5.80, p <0.01). TH cores have generally lower PC1 scores and are therefore more elongated than NK, thus supporting the elongation index based on length-to-width ratios (Table 5). TH cores tend to have higher PC3 values, indicating greater upper-surface convexity and correspondingly steep distal ridges. This difference is also expressed in higher PC5 and lower PC6 scores for TH cores, reflecting thick cores with a convex upper surface volume and a flatter lower surface. In contrast, NK cores have a more typical Levallois core morphology, with a flatter upper surface and convex lower surface. Both NK and TH encompass a similar range of variability on PC4, describing distal shape, although the highest-scoring specimens (ovate) are from the Nile Valley, and the lowest score (most pointed) is from Dhofar. It is likely that both extremes represent advanced stages of reduction.

At an assemblage level, the strongest differences are expressed on PC1 (F = 13.52) and PC3 (F = 12.17, p <0.001), with high significance still observed on PC2 (F = 5.23) and PC6 (F = 5.06, p <0.001) (Supplementary Figure S1). Most of this variation is explained by regional differences, but NK3 presents an interesting case. The core sample from NK3 lies in between the other NK and TH sites on PCs 1 to 4, showing no statistically significant differences (Supplementary Figure S1a-b, Supplementary Table S6); however, on PC6, NK3 significantly differs from NK1-M (p <0.05), TH.571 (p <0.05) and TH. 584 (p <0.001) (Supplementary Table S6). NK3 cores concentrate at the higher end of the spectrum (Supplementary Figure S1c), indicating greater lower surface volume and convexity. Based on dimensions alone, NK3 has similar mean values to the NK1 assemblages, but shows slightly higher elongation and lower flattening (Table 5). The volumetric concept underpinning Levallois cores emphasises the role of two asymmetric volumes (Boëda 1995), but this is difficult to capture quantitatively using standard metrics (Ranhorn et al. 2019). Thus, the insights offered here by comparing these aspects of shape reveal some nuanced differences between NK assemblages, as well as clear inter-regional distinctions between the Nile Valley and Dhofar (Supplementary Figure S2a-b).



#### 4.2.2. Technological variability
Core preparation strategy (using Nubian core type as a simplified indicator of scar pattern) was considered to test whether technological, rather than regional, factors drive shape variation. The strongest differences are observed on PC3 (F = 14.01) and PC4 (F = 7.62, p <0.001), with a low F-statistic but still high significance on PC1 (F = 5.44, p <0.005) (Table 8, Supplementary Figure S3). Post-hoc tests identify that these differences are between Type 1 (distally-prepared) and Type 2 (bilateral) cores on PCs 3 and 4, and between Type 1 and Type 1/2 (distal and bilateral) cores on PC3 (Supplementary Table S6). Shape variation on PC3 corresponds to the relative volume and convexity of the core surfaces, therefore Type 1 cores have a shallow upper surface convexity, whereas this convexity is steep on Type 2 cores. PC4 relates to the distal shape of the core, indicating that Type 2 cores are generally more pointed than Type 1.

There are regional differences in the principal core preparation strategy used; Type 1 dominates at NK (76% of cores), whereas TH has high proportions of both Type 2 (41%) and Type 1/2 (46%). Therefore, a MANOVA was carried out for PC3 (showing the strongest relationship) to test for interaction between region and core type. This yielded significant results, with a low F-statistic of 7.06 (Supplementary Table S7), meaning that while different core types have distinct shapes on PC3 (i.e. surface convexity), these may also vary according to region. The addition of Nubian core samples from further assemblages and regions will contribute to testing the strength of regional differences in future analyses.

#### 4.2.3. Preferential scar shape variability
The shape of the Levallois end-product was represented by the preferential (or final) scar outline on the cores. PC1 captures 52.1% of variance, with moderate contributions from PC2 (16.6%) and PC3 (11.5%) (Supplementary Table S8). PC1 represents the elongation of the scar, reflecting contrasting long, narrow (low scores) and short, wide (high scores) shapes (Figure 7). PC2 captures asymmetry in terms of left or right skew of the product distal. PC3 corresponds to whether the distal shape is convergent (high scores) or divergent (low scores), reflecting pointed or ovate shapes.

Scar shapes show strong regional differences, particularly on PC1 (F = 64.52) and PC3 (F = 22.96, p < 0.001) whereby products from TH cores tended to have higher PC3 and lower PC1 scores, representing very narrow pointed shapes (Figure 7, Table 9). NK scars showed the inverse, reflecting products with short and wide shapes that might be better classified as flakes than points Supplementary Figure S2(c-e). There were also statistically significant differences at an inter-assemblage scale (PC1: F = 17.39; PC3: F = 7.89, p < 0.001), although this is most strongly expressed as regional differences between Nile Valley and Dhofar assemblages (Table 9). However, some finer-grained patterns suggest that NK3 scar shapes are more similar to TH.584 than the other NK assemblages (Supplementary Table S10). This reinforces the observations for overall core shape, indicating that NK3 shares more similarities with TH assemblages than the other NK1 assemblages do.

Significant differences in scar shape are also present between cores of different Nubian types. As observed for core surface shape, on PC1, Type 1 core scars are significantly different from Type 2 and Type 1/2, with a post-hoc test indicating that the Type 1/2 scar pattern with the greatest (albeit weak) difference is distal-bilateral (p < 0.1) (Supplementary Table S10). A MANOVA was also carried out for scar shape on PC1, but no significant interaction between region and scar type was detected (Supplementary Table S11). This means that in terms of the elongation of preferential scar, the shape differences observed between regions and Nubian type are independent of each other.


#### 4.2.4. NK product shape variability
The shape (and size) of the final removal from a Levallois core will be sensitive to the stage and extent of core reduction, often encountering knapping accidents as the core is exhausted (e.g. step or hinge terminations). This places potential limitations on how indicative the scar shape is of products removed earlier in the core reduction. The NK assemblages contain substantial numbers of Levallois end-products, providing the opportunity to test for significant differences between last scar and product shape.

Considering first the 3D shape of product surfaces, PC1 explains 43.7% of variance, with moderate contributions from PC2 (15.0%) and PC3 (11.0%) (Supplementary Table S12). PC1 represents product elongation, with low values reflecting long and narrow shapes and high values indicating short and wide shapes (Figure 8, Supplementary Figure S2(f)). PC2 captures asymmetry in terms of left or right skew of the product distal tip. PC3 corresponds to the convergence of the laterals, with low scores reflecting triangular shapes, while high scores have a circular shape. PC4 describes whether the greatest width of the product is at the proximal (low) or the distal (high) end. The only statistically significant shape differences between assemblages is on PC3, where NK1-U differs from both NK1-M and NK3 (Supplementary Tables S13, S14). This contrasts with earlier results on cores that instead separate NK3 from the other NK assemblages, which may indicate biases either in the shapes of products taken away from the site, or in the product sample used for analysis.

A direct comparison of preferential scar and product outline shape shows significant differences on PC1 (48.9% of variance) and PC3 (12.3%) (Supplementary Table S15, Supplementary Figure S4). Differences between scar and product shape are significant in terms of artefact type only on PC1, relating to elongation (Table 10). Significant inter-assemblage differences are expressed on PC3, describing the distal shape of the Levallois product removal. This limited difference in shape suggests that core preferential scars sufficiently represent product shape variability, making this a useful proxy for studying assemblages where Levallois products are scarce, as is the case for many Nubian core assemblages (Chiotti et al. 2009; Rose et al. 2013).


### 4.3. Shape covariation and standardisation
To test the hypothesis of shape covariance – based on the principle that Levallois reduction aims to produce a product of a predetermined shape (Bordes 1980, Boëda 1988) – core outline and preferential scar shape were compared using a two-block PLS analysis. For the entire core dataset, this indicated a strong, statistically significant positive correlation between core outline and preferential scar shape (r-PLS = 0.63, p < 0.001), with 87% of covariance explained on the first dimension (Table 11). The test was then carried out for each regional sample separately to examine intra-regional patterning. For NK, a moderate correlation is observed (r-PLS = 0.46, p < 0.001), explaining 58% of covariance. In contrast, TH showed higher shape correlation, accounting for 57% of covariance, but this was not statistically significant (r-PLS = 0.56, p = 0.1), which could indicate more variability in shapes, or less distinct covariation compared to NK cores. Additional comparisons with other Dhofar assemblages could clarify this pattern regionally. For Nubian core types, moderate to strong correlations were found, with the highest significance for Type 1 cores (Table 11).

Linear regression analysis of scores for Block 1 (core outline as the predictor variable) and Block 2 (scar outline as the response variable) demonstrated that the model explained 40% of the variance for the entire core sample (Table 12, Figure 9). Separate regression analyses were also conducted on regions and core types, finding a weak linear relationship for NK cores (R2 = 0.21), indicating a less predictable correlation between core and scar shape. TH cores showed a stronger relationship (R2 = 0.31), which suggests there is less variability compared to NK. Type 2 cores displayed the strongest model fit (R2 = 0.52), indicating a more robust and predictable relationship between core and scar shape than for Type 1 (R2 = 0.27) or Type 1/2 cores (R2 = 0.35). Overall, while the two-block PLS identifies strong covariation between core outline and scar shape in general – supporting the expectation of Levallois reduction – the linear regression results suggest that core outline shape alone is not a fully predictive determinant of scar shape. This may be due to the final scars capturing a range of states of core reduction, as well as knapping errors that cause early flake termination, thus not achieving a predetermined shape.


A further test of predetermination in the Levallois reduction process involves measuring the degree of shape standardisation among the samples. Standardisation is calculated as the coefficient of variation (CV) based on the Procrustes distance from the mean shape (Muller and Clarkson 2023). The CV for overall core shape was similarly low for all assemblages with a mean of 21 (range: 20-22, Supplementary Table S18). For preferential scar shape, there was lower standardisation and more inter-assemblage variability, with the lowest and highest CV values for TH.571 (26) and TH.584 (39), while the NK sites ranged between 31 and 39 (mean: 36). The products from NK have a mean CV of 30 (range: 28-32), suggesting slightly higher standardisation in products at the sites than reflected by core removals. However, this may be influenced by the extent of core reduction, removal of products of certain shapes, or product sample selection which favoured convergent distal morphologies. It has been argued elsewhere that Nubian Levallois is a highly standardised reduction method compared with other Levallois methods (Groucutt and Rose 2023), and Dhofar in particular shows higher levels of size standardisation than Nubian cores from other regions (Samawi and Hallinan 2024). The observed covariance between core and scar shape supports a deliberate relationship in Nubian Levallois reduction, yet less standardisation in end-product shape is observed in each region. Since the final scar shape only represents the last stage of core reduction, it cannot account for the products from earlier phases – a general limitation in lithic studies (Dibble et al. 2017). Despite this, strong similarities between NK products and core scars suggest the final scar is nevertheless a useful proxy for end-product shape, allowing insights into the goals of Nubian Levallois production even when products are scarce in assemblages.

# 5. Discussion
Our analyses present a new replicable protocol for applying GM techniques to Levallois cores to study shape – a key characteristic of the Levallois concept. Using Nubian cores from two regions as a case study, we tested a series of expectations concerning different aspects of shape variability arising out of previous discussions of Levallois technology. The widely accepted definition of Levallois conceptualises the core as two hierarchically-organised surfaces, whose convexities are carefully controlled to produce a specific predetermined product shape (Boëda 1988, 1995). Boëda (2013) further introduced the characteristic of “autocorrelation”, meaning that the overall shape of Levallois cores is preserved, regardless of size due to progressive reduction. Our results support this by demonstrating low allometry at an intra-assemblage scale, even though there are marked size differences between cores regionally.

Van Peer’s (1992) prior refitting and attribute-based study of the NK assemblages found that the organisation of scar ridges (i.e. the preparation strategy) was a stronger determinant of end-product shape than core shape itself. While his work specifically compared Nubian and Classical cores, we tested variation between Nubian reduction strategies. GM analysis showed that the strongest influence of preparation (i.e. core type) was on the relative volume and convexity of the core surfaces, rather than outline shape. Type 1 (distally-prepared) cores had a shallower upper surface convexity than Type 2 (bilateral) cores. However, this pattern may be influenced by region since NK cores are predominantly Type 1 whereas TH contributes the most Type 2 cores. Other Nile Valley sites show more variation in the Nubian strategy applied, seeing higher frequencies of Type 2 and 1/2 cores (Olszewski et al. 2010; Samawi and Hallinan 2024). Based on samples from Abydos (90 km from NK), Chiotti et al. (2009) argue for fluidity between Nubian preparation strategies, observing minimal differences in size and shape that, they reason, negate differentiation between core types. Comparisons of these and other assemblages with varying proportions of distal and lateral preparation (e.g. Usik et al. 2013; Oron et al. 2024) may help to further clarify the relative role of region and scar organisation on Nubian core (and end-product) shape.

For Bordes (1950, 1980), differences in Levallois core morphology were fundamentally linked to end-product shape, serving as a key factor in distinguishing between different core types (i.e. Levallois flake, point and blade cores). Our analysis of covariation between scar and core shape confirms that this principle applies to Nubian cores specifically, revealing a strong relationship between the two. Although pointed end-products are often regarded as the primary target of Nubian reduction, GM captures a variety of preferential scar shapes (i.e. ovate to pointed), that correspond with end-product variability at NK. In prior study of the NK assemblages, Van Peer (1992) observed that Nubian end-products were more standardised than those from centripetal Levallois cores, proposing that the Nubian preparation strategy allowed for better shape control. However, Van Peer also noted that distinguishing between Nubian and centripetal end-products was not straightforward, as suggested by the presence of less pointed flake-like scar shapes on NK Nubian cores. In other Nile Valley contexts, the low frequency of pointed end-products has prompted the suggestion that pointedness may not be as critical for Nubian core products as generally assumed (Chiotti et al. 2009). Many studies observe that end-product elongation is a characteristic that differentiates Nubian from other Levallois (point) methods (Van Peer 1992; Usik et al. 2013; Hilbert et al. 2016; Oron et al. 2024; Samawi and Hallinan 2024); therefore, extending GM comparisons to include other Levallois core types could further test the distinctiveness of Nubian end-product shape.

In the only prior GM study of Levallois core morphology, Lycett and von Cramon-Taubadel (2013) predicted that the planform shape of Levallois cores was more likely to reflect spatial or temporal variations than surface topology, citing Nubian cores as one example. Our results show that the aspects of shape represented by PC1 (explaining 20% of variance) and PC2 (15%) do primarily relate to outline shape, specifically in terms of elongation (PC1) and distal skew (PC2). Only elongation showed regionally significant differences, whereby TH cores were more elongated than NK samples. However, PC1 also describes the convexity of the upper surface at its distal end – capturing a defining feature of Nubian cores, the distal median ridge. This also shows clear regional patterns on PC3 (11% of variance) which encompasses differences in relative surface volume and convexity. The Levallois concept defined by Boëda (1995) specifies that Levallois cores must possess hemispheres or surfaces with asymmetric volumes, that serve different, non-interchangeable roles during reduction. While there is a general tendency for the lower surface volume to be greater than the upper surface (e.g. for classic centripetal preferential cores), cores showing the inverse are not necessarily excluded from Levallois if all other criteria are met (maintenance of distal and lateral convexities, prepared striking platforms, exploitation parallel to the plane of intersection). Thus, while NK Nubian cores adhere very closely to this typical distribution of volumes, TH cores tend to show a steep upper (exploitation) surface with greater volume. On many TH specimens, the lower surface has very limited preparation, except for at the distal platform. One explanation for this may be that a naturally tabular shape of the raw material blocks provided the required ~80-90-degree angle for shaping of the upper face, without the need for extensive lower surface removals (Brantingham and Kuhn 2001). 

The DMR on a Nubian core serves a critical role in guiding the removal of the preferential product, and therefore in determining its (usually) pointed shape. While Van Peer’s (1992) study of the NK cores did not explicitly quantify the ridge angle, the organisation of guiding ridges, their position and inclination were key to his explanation of Levallois shape control and thus the distinction between centripetal and Nubian reduction. In their study of Dhofar assemblages (though not including TH.584 and TH.571), Usik et al. (2013) introduced measurement of the DMR steepness, grouping angles of less than 60°, 60-90° and 90-120°. They observed most specimens to fall in the steep to semi-steep (<90°) range – an observation generally supported by the few other studies that measure Nubian core ridge angles (Goder-Goldberger et al. 2016; Hilbert et al 2016; Hallinan and Shaw 2020; Oron et al. 2024). However, the Dhofar assemblages in general have a much higher incidence of steep (<60°) angles than other regions, such as South Africa (5% of measured ridges), the Negev (28-36%), and central Arabia (24-58%). DMR measurements for new Nubian core collections from the Nile Valley, near Thebes, show a mean angle of 90° (range: 60-110°) (Hallinan, unpublished data). These regional trends of relative distal ridge steepness were captured through GM, on PCs 1 and 3, indicating that Dhofar cores have a generally steeper distal ridge convexity than Nile Valley cores. The application of GM to further assemblages will be an important next step in verifying the range and extent of regional patterns in DMR angle. This approach is particularly valuable given the limitations in replicability in manual angle measurements (Dibble and Bernard 1980; Pargeter et al. 2023), and the complex adjustment parameters associated with digital measurement tools (Valletta et al. 2020; Yezzi-Woodley et al. 2021; Schunk et al. 2023; Muller et al. 2024).

Nubian cores were a suitable test case for developing our GM protocol because of their clear equivalent morphological features that can be identified across specimens. Similarly, the volumetric structure of other preferential Levallois core forms allows for the landmarking of homologous features, even in the absence of a clear distal point, provided that the orientation is consistent. To test the flexibility of our landmarking methodology on cores without a distal landmark, we re-ran our semilandmark processing and PCA analyses on the existing sample, treating only the two ends of the proximal platform as fixed landmarks. This modification resulted in a higher failure rate during the deformation of the patched SLM template over the core; seven specimens failed to deform and a further 22 experienced inversions between the patched faces. This indicates that the third landmark plays an important role in orientation and should be used where possible. Nevertheless, a substantial portion of the dataset using the modified protocol was suitable for GM analysis (83% of the original specimens), with PCA capturing the same underlying shape structure of the data, including strong regional differences. Therefore, when the specific cores under analysis are assessed carefully for potential homologous features, our method can be adapted with high success. This is particularly important in facilitating future comparisons between Nubian and other Levallois cores to assess the distinctiveness of the Nubian reduction method (cf. Blinkhorn et al. 2021b).

The advantages of 3D GM over traditional metrics in quantifying Levallois core shape have been demonstrated through this study of Nubian core variability. This is particularly evident in the shape characteristics captured by PCs 3, 5 and 6, which relate to convexity and relative surface volume – features that are difficult to represent using standard metrics. While the metric index of flattening indicates that NK cores are generally flatter than TH cores, this clearly oversimplifies the variability in core profile shape. Although GM can identify shape differences, interpreting the behavioural reasons behind them remains complex and this is not fully explored here. Factors related to raw material, technological efficiency, product function, cultural traditions and demographic relationships may all play a role and are important for future model testing (Samawi and Hallinan 2024).

# 6. Conclusion
The definition of Levallois technology has shifted from an emphasis on the outline shape of blanks produced (Bordes 1950, 1980), to the geometric configuration of the core as a three-dimensional volume (Boëda 1988, 1993, 1995). While controlling shape is a fundamental principle, a wide range of core morphologies can fall within classifications of Levallois technology, produced using different reduction strategies (Brantingham and Kuhn 2001). Nubian Levallois is a distinctive method, resulting in cores that possess specific technological and morphological features that have largely been explored through traditional metric, attribute and refitting analyses (e.g. Van Peer 1992; Chiotti et al. 2007; Usik et al. 2013; Hallinan and Shaw 2020; Oron et al. 2024; Samawi and Hallinan 2024). Our study demonstrates that 3D GM is a powerful analytical tool for quantifying shape variation, allowing the identification of nuanced patterning in regional and technological variability. Given current debate over the definition, distinctiveness, and distribution of Nubian cores (Hallinan et al. 2022a, 2022b; Hallinan 2024), further application of this method will provide important contributions to understanding variability, informing broader patterns of human adaptation and dispersal at an inter-regional scale. 

Three-dimensional scanning and methods of analysis are rapidly expanding in lithic research, especially in relation to cores which are fundamentally 3D volumes and inadequately captured by 2D measurements (e.g. Clarkson et al. 2006; Lycett et al. 2010; Shott and Trail 2010; Lycett and von Cramon-Taubadel 2013; Bretzke and Conard 2012; Porter et al. 2019; Ranhorn et al. 2019; Valletta et al. 2021; Grosman et al. 2022; Wyatt-Spratt 2022; Lombao et al. 2023; Lin et al. 2024). The importance of open-source methods has been demonstrated by the wide uptake of GM software AGMT3-D in a range of lithic studies (e.g. Herzlinger et al. 2017; Herzlinger and Grosman 2018; Delpiano and Uthmeier 2020; Shipton and White 2020; Hashemi et al. 2021; Li et al. 2021; Falcucci and Peresani 2022; Falcucci et al. 2022; García-Medrano et al. 2020, 2022). However, while AGMT3-D’s automatic orientation protocol is suitable for artefacts with certain morphologies, such as handaxes, points, bladelets and backed tools, our tests of its performance on cores, especially Nubian cores, had more limited success. GM analysis often also uses the open platform R, employing packages developed for the biological sciences, such as geomorph and Morpho (e.g. Archer et al. 2015, 2018, 2021; Schoville et al. 2023), with promising potential in development of the Lithics3D (Pop 2024). Nevertheless, the code, data and methods are rarely shared in full (although see Timbrell et al. 2022a; Hussain et al. 2023; Selden 2023; Thulman et al. 2023 (2D GM); Bustos-Pérez et al. 2024 (3D GM)). The methodology presented here, supported by fully-available R script and data, aims to meet the increasing demand for replicable and reproducible methods in archaeology, particularly lithic studies (e.g. Marwick 2017; Karoune and Plomp 2022; Timbrell et al. 2022b; Araujo et al. 2023; Di Maida et al. 2023; Pargeter et al. 2023), offering a novel application of 3D GM to Levallois technology that can be adapted to other types of cores with highly structured geometries. 



# Acknowledgements
We are grateful to Philip Van Peer (Nazlet Khater) and Jeffrey Rose (Dhofar) for study permission and facilitating access to the assemblages.

# Funding
This research was funded by the European Union Horizon 2020 Framework under the Marie Skłodowska‐Curie Grant Agreement No. 891917, awarded to Emily Hallinan, hosted by João Cascalheira (project TANKwA). Emily Hallinan is currently supported by grant #2021/00041.CEECIND/CP1672/CT0007, funded by the Portuguese Foundation for Science and Technology (FCT, Fundação para a Ciência e Technologia) (project NuBIAN).


# References





\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
