---
title: "GM alternative protocol"
author: "Emily Hallinan"
date: "2024-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, echo = TRUE, results = 'hide', message = FALSE, warning = FALSE, fig.show = 'hide')
```

------------------------------------------------------------------------

## 5A. Alternative method for preparing core data without a distal landmark

**Software:** R (R Core Team 2024)\
**Packages:**\
- Morpho (Schlager 2017) (v2.12)\
- Rvcg (Schlager 2017) (v0.23)\
- geomorph (Baken et al. 2021; Adams et al. 2024) (v4.0.8)\
- rgl (Murdoch & Adler 2024) (v1.3.1)\
- dplyr (Wickham et al. 2023) (v1.1.4)\
- abind (Plate & Heiberger 2016) (v1.4.5)\

```{r Step 5A setup}
# Install and load required packages
library(Morpho)
library(Rvcg)
library(geomorph)
library(rgl)
library(dplyr)
library(ggplot2)
library(abind)
```

**Data:** Download, from our OSF.io repository the corresponding .ply and .pts files for 115 cores from NK sites and 51 from TH sites (total, 166).\

```{r Step 5 DownloadCores}

## Core Data
# Create the directory to store core data if it doesn't exist
core_data <- file.path("data", "raw_data", "core_data")
dir.create(core_data, recursive = TRUE, showWarnings = FALSE)

# Download core data from OSF (alternatively, use the link in a web browser to download the .zip file)
osf_retrieve_file("https://osf.io/mt4xv") %>%
  osf_download(path = "data/raw_data/core_data")

# Specify the path to the Zip file
zip_file <- "data/raw_data/core_data/core_data.zip"

# Extract files from the Zip archive
unzip(zip_file, exdir = "data/raw_data/core_data")

```

**Method:**

1.  The corresponding 3D models (.ply) and landmark co-ordinates (.pts) for points 2 to 63 on each specimen were imported into R in a single matrix, then converted into a data array (package: geomorph, function: arrayspecs). This excludes the distal landmark in the original .pts file.

```{r Step 5A.1}

# Create list of core names
corelist <- list.files(path = core_data, pattern = ".pts")

# Get names from files
names <- gsub(".pts", "", basename(corelist))

# Create empty array
nd.core_landmarks <- NULL

# Import each file (selecting landmarks 2 to 63) (i.e. not the distal LM) and paste them in a single matrix
for (i in 1:length(corelist)){
  core_path <- file.path(core_data, corelist[i])
  tmp <- as.data.frame(read.table(core_path, skip = 2, header = F)[2:63,2:4])
  tmp <- as.matrix(tmp)
  nd.core_landmarks <- rbind(nd.core_landmarks, tmp)
}

nd.core_landmarks <- mapply(nd.core_landmarks, FUN=as.numeric)
nd.core_landmarks <- matrix(data=nd.core_landmarks, ncol=3)

# Convert landmark 3D data matrix into array (62 lms, 3 dimensions)
nd.core_landmarks <- geomorph::arrayspecs(nd.core_landmarks, 62, 3)

# View specimen example with imported landmark configuration
core1 <- vcgImport(file.path(core_data, "CoreME78.627Aa_LR.ply"))
shade3d(core1, color = "gray")
spheres3d(nd.core_landmarks[1:2,,34], color = 1, radius = 1)
spheres3d(nd.core_landmarks[3:22,,34], color = 2, radius = 0.7)
spheres3d(nd.core_landmarks[23:62,,34], color = 3, radius = 0.7)
#rgl.snapshot('nd.core_landmarks_example.png', fmt = 'png')
close3d()
```

2.  **Core outline:** A new array was created for resampled SLMs, with dimensions k x m x n (38 points (because no distal landmark), 3 x-y-z co-ordinate dimensions, 166 specimens). SLMs were resampled using only two curves (platform and core body outline) and redistributed with equal spacing on all specimens (package: geomorph, function: digit.curves), formed of 6 platform points, and one edge curves of 30 points.

```{r Step 5A.2}
# Create a new data array for resampled semilandmarks (38 slms, 3 dimensions, 166 specimens)
nd.core_lms <- array(0, dim=c(38,3,166))

# Resample semilandmarks on curves between fixed landmarks and redistribute to be equally-spaced
for(i in 1:dim(nd.core_landmarks)[3]) {
  startA <- nd.core_landmarks[2,,i]
  startB <- nd.core_landmarks[1,,i]
  curveA <- digit.curves(startA, rbind(nd.core_landmarks[3:22,,i], startB), nPoints = 6, closed = F)
  curveB <- digit.curves(startB, rbind(nd.core_landmarks[23:62,,i], startA), nPoints = 30, closed = F)
  comb <- rbind(startA, startB, curveA[2:(nrow(curveA)-1),], curveB[2:(nrow(curveB)-1),])
  nd.core_lms[,,i] <- comb
}

# Save the new semilandmark array for ease of future use
save(nd.core_lms, file = "../methods/data/derived_data/nd.core_lms.RData")

# To view an example core mesh (.ply) with resampled slms (identify the correct specimen number (#34) by viewing 'corelist')
shade3d(core1, color = "gray")
spheres3d(nd.core_lms[1:2,,34], color = 1, radius = 1)
spheres3d(nd.core_lms[3:8,,34], color = 2, radius = 0.7)
spheres3d(nd.core_lms[9:38,,34], color = 3, radius = 0.7)
#rgl.snapshot('nd.core_lms_example.png', fmt = 'png')
close3d()
```

3.  The resampled SLMs were then slid (relaxed) along the core edge to mimimise the Procrustes distance across the dataset using the Procrustes consensus (mean shape) as a reference (package: Morpho, function: slider3d). Two fixed landmark points (“fix”) were defined (points 1-2), and two curves (“outlines”) formed by sliding points (points 3-8 and 10-38). The new SLM points ($dataslide) were saved in a new data array.

```{r Step 5A.3}
# Define fixed LMs (that do not slide) and semilandmarks on outline curves to slide
fix <- c(1:2)
outlineA <- c(3:8)
outlineB <- c(9:38)

outlines <- list(outlineA, outlineB)

dimnames(nd.core_lms)[[3]] <- names

# Slide semilandmarks along curves (this step takes some time and processing power for the large sample size)
slide_curves <- slider3d(nd.core_lms, SMvector = fix, deselect = TRUE, outlines = outlines,
                         sur.path = "../methods/data/raw_data/core_data", sur.type = "ply", iterations = 3)

# Visualise slid slms from previous to new positions
deformGrid3d(slide_curves$dataslide[,,34],nd.core_lms[,,34],ngrid = 0)

# View slid slms on mesh model
shade3d(core1, color = "gray")
spheres3d(slide_curves$dataslide[fix,,34],col=1,radius=1)
spheres3d(slide_curves$dataslide[,,34],col=3,radius=0.7)
#rgl.snapshot('nd.core_lms_slid.png', fmt = 'png')
close3d()

# Rename and save outline slms
nd.alloutline <- slide_curves$dataslide
save(nd.alloutline, file = "../methods/data/derived_data/nd.alloutline.RData")
```

4.  **Preferential scar outline:** (This process is unaltered). A new array was created (“pref_landmarks”) and the 50 preferential scar outline landmarks (points 71 to 120) imported for each specimen. The SLM resampling (digit.curves) and sliding process (silder3d) was repeated for the preferential scar outline, with two curves formed of 30 edge points and 5 platform points, saved in a new data array “allprefs”, with dimensions 37 x 3 x 166. Note that because the sliding process allows SLMs to be slid along the core model surface, which here extends beyond the limits of the preferential scar outline, the platform SLMs (31-38) were treated as fixed points and excluded from the second relaxation step to avoid their misplacement.

```{r Step 5A.4}
# Create a new data array for pref scar landmarks
pref_landmarks <- NULL

# Import each file and paste them in a single matrix (this extracts only landmarks 71 to 120 from the .pts file)
for (i in 1:length(corelist)){
  core_path <- file.path(core_data, corelist[i])
  tmp <- as.data.frame(read.table(core_path, skip = 2, header = F)[71:120,2:4])
  tmp <- as.matrix(tmp)
  pref_landmarks <- rbind(pref_landmarks, tmp)
}

pref_landmarks <- mapply(pref_landmarks, FUN=as.numeric)
pref_landmarks <- matrix(data=pref_landmarks, ncol=3)

# Convert landmark 3D data matrix into array (50 lms, 3 dimensions)
pref_landmarks <- geomorph::arrayspecs(pref_landmarks, 50, 3)

# View specimen example with imported landmark configuration
shade3d(core1, color = "gray")
spheres3d(pref_landmarks[1,,34], color = 1, radius = 1)
spheres3d(pref_landmarks[40,,34], color = 1, radius = 1)
spheres3d(pref_landmarks[2:39,,34], color = 3, radius = 0.7)
spheres3d(pref_landmarks[41:50,,34], color = 4, radius = 0.7)
#rgl.snapshot('pref_landmarks_example.png', fmt = 'png')
close3d()

# Create an empty array for resampled semilandmarks (37 slms, 3 dimensions, 166 specimens)
pref_lms <- array(0, dim=c(37,3,166))

# Redistribute equally-spaced semilandmarks on all specimens along two curves (5 platform and 30 edge points)
for(i in 1:dim(pref_landmarks)[3]) {
  startA <- pref_landmarks[1,,i]
  startB <- pref_landmarks[40,,i]
  curveA <- digit.curves(startA,rbind(pref_landmarks[2:39,,i], startB), nPoints = 30, closed = F)
  curveB <- digit.curves(startB, rbind(pref_landmarks[41:50,,i], startA), nPoints = 5, closed = F)
  comb <- rbind(curveA[2:nrow(curveA),], curveB[2:nrow(curveB),])
  pref_lms[,,i] <- comb
}

shade3d(core1, color = "gray")
spheres3d(pref_lms[37,,34], color = 1, radius = 1)
spheres3d(pref_lms[31,,34], color = 1, radius = 1)
spheres3d(pref_lms[1:30,,34], color = 3, radius = 0.7)
spheres3d(pref_lms[32:36,,34], color = 4, radius = 0.7)
#rgl.snapshot('pref_lms_example.png', fmt = 'png')
close3d()

# Save for future use
save(pref_lms, file = "../methods/data/derived_data/pref_lms.RData")

# Fix platform landmarks and relax edge semilandmarks on pref scar outline
fixP <- c(31:37)
outlinesP <- c(1:30)

dimnames(pref_lms)[[3]] <- names

# Slide semilandmarks along curves (this step takes some time and processing power for the large sample size)
slide_curvesP <- slider3d(pref_lms, SMvector = fixP, deselect = TRUE, outlines = outlinesP,
                          sur.path = "../methods/data/raw_data/core_data", sur.type = "ply", iterations = 3)

# Visualise slid slms from previous to new positions
deformGrid3d(slide_curvesP$dataslide[,,34],pref_lms[,,34],ngrid = 0)

# View slid slms on mesh model
shade3d(core1, color = "gray")
spheres3d(slide_curvesP$dataslide[fixP,,34],col=4,radius=1)
spheres3d(slide_curvesP$dataslide[,,34],col=3,radius=0.7)
#rgl.snapshot('pref_lms_slid.png', fmt = 'png')
close3d()

# Rename and save
allprefs <- slide_curvesP$dataslide
save(allprefs, file = "../methods/data/derived_data/allprefs.RData")
```

5.  **Core surface patch:** (Resume alterations here). A template artefact was selected (CoreME78.389c) that did not represent any extremes in artefact shape, and two surface patches were placed (see 4.5 above). An atlas was created from the template mesh (package: Morpho, function: createAtlas) with the core outline defined by the slid SLM points.

```{r Step 5A.5}
# Create atlas using (arbitrarily selected) core ME78.389c (specimen #20 for slid outline slms) as a template
coreT <- vcgImport(file.path(core_data, "CoreME78.389c_LR.ply"))
core_temp <- as.matrix(read.table("../methods/data/derived_data/CoreME78.389c_LRtemplate.pts", skip = 2, header = F)[,2:4])

nd.core_atlas <- createAtlas(coreT, landmarks = nd.alloutline[,,20], patch = core_temp)
plotAtlas(nd.core_atlas)
```

6.  The surface points (patch) from the template (atlas) were warped and projected onto the shape of each specimen (function: placePatch), placing 400 surface SLMs, saved as a new data array combining outline and surface points, with dimensions 438 x 3 x 166.

```{r Step 5A.6}
# Place the patch onto all specimens. This deforms the template over each mesh surface. The inflate parameter can be adjusted if there are placement errors observed in the checking stage
nd.allpatched <- placePatch(atlas = nd.core_atlas, dat.array = nd.alloutline, path = "../methods/data/raw_data/core_data", inflate = 10, fileext = ".ply")

# Save the patched slm data
save(nd.allpatched, file = "../methods/data/derived_data/nd.allpatched.RData")
```

7.  The deformation was checked for each specimen (function: checkLM), visualised in 3D using the Morpho dependent package rgl. Where deformation was not successful (i.e. visualisation did not show the expected placement of two surface patches (green spheres) separated by a semilandmarked outline (red spheres); or there were clear gaps in the patch grid), these specimens were extracted to attempt deformation again with an adjusted inflate parameter (here, 15 gave improved results).

```{r Step 5A.7a}
# Check patch slms have deformed correctly on each specimen (this is time consuming but essential to check the deformation process).
checkLM(nd.allpatched, atlas=nd.core_atlas)

## Problems in deformation identified in specimens 53, 59, 67, 82, 88, 104, 128 (minor) and 87, 99, 125, 133, 134 (major)

# Separate the failed specimens
failed <- c(53, 59, 67, 82, 88, 104, 128, 87, 99, 125, 133, 134)

# Separate failed and successful specimens
exclude <- rep(FALSE, 166)
exclude[failed] <- TRUE

outline_failed <- nd.alloutline[,,exclude]
patched_good <- nd.allpatched[,,!exclude]
```

```{r Step 5A.7b}
# Re-run the patch deformation on the failed specimens with a higher inflate parameter
nd.patched_new <- placePatch(atlas = nd.core_atlas, dat.array = outline_failed, path = "../methods/data/raw_data/core_data", inflate = 15, fileext = ".ply")
checkLM(nd.patched_new, atlas=nd.core_atlas)

## Problems still identified in specimens 2, 3, 5, 7, 9, 11, 12 - repeat the exclusion process
# Separate the failed specimens
failed <- c(2, 3, 5, 7, 9, 11, 12)

# Separate failed and successful specimens
exclude <- rep(FALSE, 12)
exclude[failed] <- TRUE
patched_new2 <- nd.patched_new[,,!exclude]

# Create a new array and combine both sets of successfully patched specimens on the 3rd dimension
nd.allpatched <- array(NA, dim = c(dim(patched_good)[1], dim(patched_new2)[2], 159))
nd.allpatched <- abind(patched_good, patched_new2, along = 3)

# Ensure the dimnames are preserved
dimnames(nd.allpatched)[[3]] <- c(dimnames(patched_good)[[3]], dimnames(patched_new2)[[3]])
```

 8.  Owing to increased patch deformation failures, further checks for errors were conducted. A GPA was carried out, then a looped process plotted the Procrustes-aligned landmark configurations and output each as an image so that they could be visually inspected. This allowed specimens with inverted faces (i.e. the upper and lower surface patches deformed to the wrong surface) to be isolated. Inverted specimens will also be apparent as outliers when plotted on a PCA.
 
```{r Step 5A.8b}
# Check for errors in alignment (e.g. inversions) by conducting GPA and inspecting landmark configurations.
nd.patched_gpa <- gpagen(nd.allpatched,PrinAxes=TRUE)

# Extract the coordinate data from the gpa
coords <- nd.patched_gpa$coords

# Get the number of specimens
num_specimens <- dim(coords)[3]

# Loop through each specimen and create a separate plot, then loop through each landmark to plot configurations in different colours. These are saved individually.
for (specimen_idx in 1:num_specimens) {
  specimen_coords <- coords[, , specimen_idx]
  for (landmark_idx in 1:438) {
    landmark <- specimen_coords[landmark_idx, , drop = FALSE]
    x <- as.numeric(landmark[, 1]) # Extract x,y,z coordinates
    y <- as.numeric(landmark[, 2])
    z <- as.numeric(landmark[, 3])
    if (landmark_idx %in% c(1:2)) {
      col <- "red"  # Fixed LM1-2 in red
    } else if (landmark_idx %in% c(3:38)) {
      col <- "blue"  # Outline LMs3-38 in blue
    } else if (landmark_idx %in% c(39:238)) {
      col <- "green"  # Upper surface patch in green
    } else {
      col <- "yellow"  # Lower surface patch in yellow
    }
    points3d(x, y, z, col = col, size = 5)
  }
  view3d(theta = 180, phi = 0, zoom = 1, fov = 0, interactive = TRUE)
  rgl.snapshot(file = paste("specimen_", specimen_idx, ".png", sep = ""), fmt = "png")
  close3d()
}

## 22 specimens show inversions: 10, 14, 31, 35, 37, 38, 62, 67, 69, 73, 76, 78, 80, 97, 104, 111, 143, 145, 155, 156, 159

# Identify and exclude the failed specimen numbers
failed <- c(10, 14, 31, 35, 37, 38, 62, 67, 69, 73, 76, 78, 80, 97, 104, 111, 143, 145, 155, 156, 159)
nd.filtered <- nd.allpatched[, , -failed]

# Repeat GPA
nd.patched_gpa <- gpagen(nd.filtered,PrinAxes=TRUE)


### Modified core data are now ready for GM analyses

## R data files can be loaded as:
load("../methods/data/derived_data/nd.allpatched.RData")
load("../methods/data/derived_data/nd.alloutline.RData")
```

9.  Outlier shapes (i.e. those with extreme values not necessarily caused by errors) were excluded based on Procrustes distance from the mean, using a threshold of 3 standard deviations.

```{r Step 5A.9}
## STEP 5A.9 Check for outliers using Procrustes distances
# Calculate Procrustes distances to the mean shape
meansh <- mshape(nd.patched_gpa$coords)
proc_dist <- apply(nd.patched_gpa$coords, 3, function(x) sqrt(sum((x - meansh)^2)))

# Calculate the mean and standard deviation of the Procrustes distances
mean_proc <- mean(proc_dist)
sd_proc <- sd(proc_dist)

# Define the threshold for outliers within 3 sd of the mean (i.e. 99.7% of the data)
threshold_upper <- mean_proc + 3 * sd_proc
threshold_lower <- mean_proc - 3 * sd_proc

# Identify outliers
outliers <- proc_dist < threshold_lower | proc_dist > threshold_upper

# Display outlier specimens
outlier_specs <- proc_dist[outliers]
print(outlier_specs)

# Exclude outlier spec (CoreME78.373.37) from the (filtered) patched core dataset and rerun GPA on new dataset
outlier_specs <- 56
all_specs <- 1:138
keep <- setdiff(all_specs, outlier_specs)
nd.filtered <- nd.filtered[, , keep]
nd.patched_gpa <- gpagen(nd.filtered,PrinAxes=TRUE) # check this has the right no of specimens
```

10.  A Principal Components Analysis (PCA) was carried out on the Procrustes co-ordinates for the GPA dataset (package: geomorph, function: gm.prcomp). The object generated contains the PC scores (x) which were extracted and saved in a new dataframe with attribute data for each specimen appended. The eigenvalues (d) were also extracted to a table to show the amount of variance each component accounts for in the dataset. PCA plots were generated using ggplot2.

```{r Step 5A.10a}
## STEP 5A.10 Conduct PCA
# Run PCA and extract PC scores to new dataframe
PCA_allpatch.nd <- gm.prcomp(nd.patched_gpa$coords)
PCscores_patch.nd <- PCA_allpatch.nd$x # x represents the PC scores
coord_allpatch.nd <- as.data.frame(PCscores_patch.nd)

# Import attribute files
cores <- read.csv("../methods/data/derived_data/File_list_cores.csv")
cores <- dplyr::arrange(cores, File_name)

# Match the coord to the core attribute data only for the specimens remaining in the sample
coord_allpatch.nd$File_name <- rownames(coord_allpatch.nd)
coord_allpatch.nd <- merge(coord_allpatch.nd, cores, by = "File_name")


# Generate variance tables that show the percentage of variation captured by each component
eigenvalues <- PCA_allpatch.nd$d # d represents the eigenvalues
pc_numbers <- 1:length(eigenvalues)
perc_variance <- (eigenvalues / sum(eigenvalues)) * 100
cum_perc <- cumsum(perc_variance)
perc_table_patch.nd <- data.frame(PC = 1:length(eigenvalues), Eigenvalue = eigenvalues, PercVariance = perc_variance, CumulativePerc = cum_perc)
```

```{r Step 5A.10b}
# Generate PCA plots (here with aesthetics set for area/region)
# Set colour/shape aesthetics for the NK and TH samples for the plots
colsArea <- c("NK" = "royalblue", "TH" = "indianred1")
shapesArea <- c("NK" = 16, "TH" = 17)

#PC1 and 2
coord_allpatch.nd %>%
  ggplot(aes(x = Comp1, y = Comp2, shape = Area)) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  geom_point(aes(color = Area)) +
  scale_color_manual(values = colsArea) +
  scale_fill_manual(values = colsArea) +
  stat_ellipse(geom = "polygon", aes(fill = Area), alpha = 0.2, show.legend = TRUE, level = 0.95) +
  labs(shape = "Region", color = "Region", fill = "Region") +
  labs(x = "PC1", y = "PC2") +
  theme_minimal() +
  theme(legend.position = "right") +
  ggtitle("Core surfaces: PCA (PC1-PC2)")

#PC3 and 4
#CoreME78.737.37_LR (Spec 56) exhibits anomalous values on PC3.
coord_allpatch.nd %>%
  ggplot(aes(x = Comp3, y = Comp4, shape = Area)) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  geom_point(aes(color = Area)) +
  scale_color_manual(values = colsArea) +
  scale_fill_manual(values = colsArea) +
  stat_ellipse(geom = "polygon", aes(fill = Area), alpha = 0.2, show.legend = TRUE, level = 0.95) +
  labs(shape = "Region", color = "Region", fill = "Region") +
  labs(x = "PC3", y = "PC4") +
  theme_minimal() +
  theme(legend.position = "right") +
  ggtitle("Core surfaces: PCA (PC3-PC4)")

#PC5 and 6
coord_allpatch.nd %>%
  ggplot(aes(x = Comp5, y = Comp6, shape = Area)) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  geom_point(aes(color = Area)) +
  scale_color_manual(values = colsArea) +
  scale_fill_manual(values = colsArea) +
  stat_ellipse(geom = "polygon", aes(fill = Area), alpha = 0.2, show.legend = TRUE, level = 0.95) +
  labs(shape = "Region", color = "Region", fill = "Region") +
  labs(x = "PC5", y = "PC6") +
  theme_minimal() +
  theme(legend.position = "right") +
  ggtitle("Core surfaces: PCA (PC5-PC6)")
```

11.  Mean, minimum and maximum shapes were visualised to understand how shape varied on each PC. Further GM analysis and statistical tests can proceed as described in Step 8 of the original method.

```{r Step 5A.11}
# Find the theoretical mean shape
msh_patch <- mshape(nd.patched_gpa$coords)

# Create an outline by linking consecutive slms of the shape
#define.links(msh_patch, ptsize = 3, links = NULL) # can define links between slms manually or input as a .csv
core_links.nd <- read.csv("../methods/data/derived_data/Core_links.nd.csv")

# Identify the specimen that most closely matches the theoretical mean shape
findMeanSpec(nd.patched_gpa$coords)
#CoreME78.778d_LR 
#64 

# Define a function to generate and save min./max. shape snapshots for a given PC
GeneratePCShapes <- function(pc_num) {
  # Mean shape (grey) and minimum shape (blue)
  plotRefToTarget(msh_patch, PCA_allpatch.nd$shapes[[paste0("shapes.comp", pc_num)]]$min,
                  method = "points", links = core_links.nd,
                  gridPars = gridPar(pt.bg = "grey", pt.size = 0.5, link.lwd = 2,
                                     tar.pt.bg = "blue", tar.pt.size = 0.7,
                                     tar.link.col = "blue", tar.link.lwd = 2))
  
  # Save output in multiple orientations
  view3d(theta = 180, phi = -90, zoom = 0.8, fov = 0, interactive = TRUE)
  rgl.snapshot(paste0("PC", pc_num, "_min_profile.png"), fmt = 'png')
  view3d(theta = 180, phi = 0, zoom = 0.8, fov = 0, interactive = TRUE)
  rgl.snapshot(paste0("PC", pc_num, "_min.png"), fmt = 'png')
  view3d(theta = 90, phi = 0, zoom = 0.8, fov = 0, interactive = TRUE)
  rgl.snapshot(paste0("PC", pc_num, "_min_dist.png"), fmt = 'png')
  
  # Mean shape (grey) and maximum shape (red)
  plotRefToTarget(msh_patch, PCA_allpatch.nd$shapes[[paste0("shapes.comp", pc_num)]]$max,
                  method = "points", links = core_links.nd,
                  gridPars = gridPar(pt.bg = "grey", pt.size = 0.5, link.lwd = 2,
                                     tar.pt.bg = "red", tar.pt.size = 0.7,
                                     tar.link.col = "red", tar.link.lwd = 2))
  
  # Save output in multiple orientations
  view3d(theta = 180, phi = -90, zoom = 0.8, fov = 0, interactive = TRUE)
  rgl.snapshot(paste0("PC", pc_num, "_max_profile.png"), fmt = 'png')
  view3d(theta = 180, phi = 0, zoom = 0.8, fov = 0, interactive = TRUE)
  rgl.snapshot(paste0("PC", pc_num, "_max.png"), fmt = 'png')
  view3d(theta = 90, phi = 0, zoom = 0.8, fov = 0, interactive = TRUE)
  rgl.snapshot(paste0("PC", pc_num, "_max_dist.png"), fmt = 'png')
}

# Loop through PCs 1 to 6 (or desired number of PCs)
for (pc in 1:6) {
  GeneratePCShapes(pc)
}

# Identify the specimens corresponding to the minimum and maximum shapes
# Create a new dataframe to hold the extracted specimen information
Specs <- data.frame(PC = character(), MinSp = character(), MaxSp = character(), stringsAsFactors = FALSE)

# Identify specimen names in coord_allpatch dataframe
coord_allpatch.nd <- data.frame(
  Specimen = coord_allpatch.nd$File_name,
  coord_allpatch.nd,
  stringsAsFactors = FALSE)

# Extract specimen names that correspond to min. and max. value for first 6 PCs
for (i in 1:6) {
  pc_name <- paste0("Comp", i)
  min_specimen <- coord_allpatch.nd$Specimen[which.min(coord_allpatch.nd[[pc_name]])]
  max_specimen <- coord_allpatch.nd$Specimen[which.max(coord_allpatch.nd[[pc_name]])]
  Specs <- rbind(Specs, data.frame(PC = pc_name, MinSp = min_specimen, MaxSp = max_specimen, stringsAsFactors = FALSE))
}
print(Specs)

## Continue with other STEP 8 GM analyses as needed.
```
